ODGI: understanding pangenome graphs

### 2022-02-28 manuscript/fig_evaluation.tex
unreadme metrics section

\begin{figure}[]
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include first image
		\includegraphics[width=1.0\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/build_eval.pdf}
		\label{fig:eval-build}
	\end{subfigure}
	\begin{subfigure}{1\linewidth}
		\caption{}
		\centering
		% include fourth image
		\includegraphics[width=\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/viz_eval.pdf}
		\label{fig:eval-viz}
	\end{subfigure}
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include second image
		\includegraphics[width=\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/extract_eval.pdf}
		\label{fig:eval-extract}
	\end{subfigure}
	\caption{TODO}
	\label{fig:evaluation}
\end{figure}

>>>>

\begin{figure}[]
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include first image
		\includegraphics[width=1.0\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/build_eval.pdf}
		\label{fig:eval-build}
	\end{subfigure}
	\begin{subfigure}{\linewidth}
	\caption{}
	\centering
	% include second image
	\includegraphics[width=\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/extract_eval.pdf}
	\label{fig:eval-extract}
	\end{subfigure}
	\begin{subfigure}{1\linewidth}
		\caption{}
		\centering
		% include fourth image
		\includegraphics[width=\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/viz_eval.pdf}
		\label{fig:eval-viz}
	\end{subfigure}
	\caption{\REVIEWED{Performance on a graph of human chromosome 6 from the HPRC. ODGI compares favorably to VG across all routine pangenomic tasks. Evaluations across threads were done using a 64 human haplotype graph. Evaluations across haplotypes were done using 16 threads. \textbf{(a)} Performance evaluation when translating a graph into the tools' respective native formats. \textbf{(b)} Performance evaluation when extracting the centromeric region from the HPRC graph. \textbf{(c)} Performance evaluation when visualizing a graph. Both tools were run with only one thread. \textit{vg viz}: \textbf{*}A 816MB SVG was produced which can't be opened by any program. \textbf{**}All produced SVGs only contain an XML header, nothing else.}}
	\label{fig:evaluation}
\end{figure}

### 2022-02-28 manuscript/fig_metrics.tex
unreadme metrics section

\begin{figure}[ht!]
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include first image
        \includegraphics[width=1.0\linewidth, trim=-2.25cm 1.0cm 0cm 3.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats_svg}
        \label{fig:metrics-multiqc}
        	\vspace{-2em}
    \end{subfigure}
    \begin{subfigure}{1\linewidth}
        \caption{}
        \centering
        % include fourth image
        \includegraphics[width=\linewidth,trim=-.225cm 4cm +.425cm +3cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
        \label{fig:metrics-degree}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include second image
        \includegraphics[width=\linewidth,trim=-.225cm 3.3cm +0.425cm +3cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
        \label{fig:metrics-depth}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include second image
        \includegraphics[width=1.0\linewidth, trim=-1.75cm 2.2cm -0.75cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg.pdf}
        \label{fig:metrics-viz}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include second image
        \includegraphics[width=1.0\linewidth, trim=-0.4cm 0.4cm 3.15cm 0cm]{fig/metrics/chr4_pan_HTTex1_STR_xg_svg}
        \label{fig:metrics-str}
        \vspace{-0.5em}
    \end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
	\caption{Features of a 90-haplotype human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module. The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \citep{Piovesan2019} is in accordance with the literature \citep{Neueder2017}. \textbf{(b)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph. Around position 200 there is a huge variation in node degree. \textbf{(c)} Per nucleotide node depth distribution of CHM13 in the \textit{HTTexon1} graph. The alternating depth around position 200 indicates polymorphic variation complementing the above node degree analysis. \textbf{(d)} \textit{odgi viz} visualization of the 23 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph. \textbf{(e)} \textit{vg viz} nucleotide-level visualization of 10 gene alleles, CHM13, GRCH38 of the \textit{HTTexon1} graph focusing on the CAG variable repeat region. Figures \textbf{(b)-(e)} highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \citep{Nance1999}.}
	\label{fig:metrics}
\end{figure}

>>>>

\begin{figure}[]
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include first image
        \includegraphics[width=1.0\linewidth, trim=-2.25cm 1.0cm 0cm 3.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats_svg}
        \label{fig:metrics-multiqc}
        	\vspace{-2em}
    \end{subfigure}
    \begin{subfigure}{1\linewidth}
        \caption{}
        \centering
        % include fourth image
        \includegraphics[width=\linewidth,trim=-.225cm 4cm +.425cm +3cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
        \label{fig:metrics-degree}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include second image
        \includegraphics[width=\linewidth,trim=-.225cm 3.3cm +0.425cm +3cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
        \label{fig:metrics-depth}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include second image
        \includegraphics[width=1.0\linewidth, trim=-1.75cm 2.2cm -0.75cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg.pdf}
        \label{fig:metrics-viz}
    \end{subfigure}
    \begin{subfigure}{\linewidth}
        \caption{}
        \centering
        % include second image
        \includegraphics[width=1.0\linewidth, trim=-0.4cm 0.4cm 3.15cm 0cm]{fig/metrics/chr4_pan_HTTex1_STR_xg_svg}
        \label{fig:metrics-str}
        \vspace{-0.5em}
    \end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
	\caption{Features of a 90-haplotype human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module. \textbf{(b)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph. Around position 200 there is a huge variation in node degree. \textbf{(c)} Per nucleotide node depth distribution of CHM13 in the \textit{HTTexon1} graph. The alternating depth around position 200 indicates polymorphic variation complementing the above node degree analysis. \textbf{(d)} \textit{odgi viz} visualization of the 23 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph. \textbf{(e)} \textit{vg viz} nucleotide-level visualization of 10 gene alleles, CHM13, GRCH38 of the \textit{HTTexon1} graph focusing on the CAG variable repeat region.}
	\label{fig:metrics}
\end{figure}

### 2022-02-28 manuscript/main.tex
minimal performance figure caption
unreadme metrics section

\begin{itemize}
\item~\textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
\item~\textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
\item~\textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item~\textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item~\textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:extract}).
\item~\textit{odgi explode} breaks the graph into connected components (\S\ref{sec:extract}).
\item~\textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:extract}).
\item~\textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
\item~\textit{odgi unchop} combines unitig nodes (\S\ref{sec:edit}).
\item~\textit{odgi break} removes cycles in the graph (\S\ref{sec:edit}).
\item~\textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
\item~\textit{odgi groom} resolves spurious inverting links (\S\ref{sec:edit}).
\item~\textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).%
%\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
\item~\textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item~\textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
\item~\textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item~\textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item~\textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:metrics}).
\item~\textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:metrics}).
\item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:metrics}).
\item~\textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item~\textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
\item~\textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item~\textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

@@

The ODGI data structure (Algorithm~\ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA file.
Then, we serially add edges in the $Node.edges$ vectors of pairs of nodes.
Finally, we create paths in serial, and extend them in parallel by obtaining the mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
This parallelism speeds ODGI model construction by many-fold when testing against graphs made from assemblies produced by the HPRC (\autoref{Evaluation}).

@@

Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
To investigate on the \REVIEWED{[1,2] genomic sequences [1,2]} encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
\textit{odgi flatten} generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.

Applying \textit{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
It outputs pangenome statistics in tab-separated values (TSV) or YAML textual file formats.
MultiQC's~\citep{Ewels_2016} ODGI module %\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)}
provides an interactive way to comparatively explore such statistics of an arbitrary number of graphs.

ODGI also offers more advanced tools for the interrogation of the graphs.
To study very large pangenomes, users can use \textit{odgi bin} to summarize the path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
%This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. %maybe we want to save this up for the discussion

Genomes presenting sequences with highly identical repeats result in pangenome graphs with complex motifs that can be detected by \textit{odgi depth} and \textit{odgi degree}, which return the node depth and node degree, respectively, as defined by query criteria.
Both tools emit the output in BED format, allowing users to assess the complexity of the graph and detect intricate regions.
Indeed, high depth/degree nodes can be the mirror of genetic variation (Fig.~\ref{fig:metrics}), but also misassemblies or problems in the pangenome building, making the tools further useful for graph validation.
%In Fig.~\ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene~\citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}) are shown.

@@

In the following an example which is based on the graph in Figure~\ref{fig:jaccard}.
\input{fig_jaccard}
Let's assume, we want to translate the query path position \textit{query:3} to a path position in path \textit{target} with a context distance of 2 nucleotides per direction. Node \textit{2} is where \textit{query:3} is located. The node is traversed two times by \textit{target}. Therefore, we apply the \textit{path jaccard} concept to do a precise position translation. Starting from node \textit{2}, we collect nodes $\{1,2,4\}$ for path \textit{query}. For path \textit{target} we have two possible steps to start our collection from: The first or the second step in node \textit{2}. We obtain the node sets $T_1=\{1,2,2\}$ and $T_2=\{2,2,3\}$.
\begin{multline}
	J_1(T_1,Q)=\frac{|T_1\cap Q|}{|T_1\cup Q|}=\frac{|\{1,2,2\}\cap \{1,2,4\}|}{|\{1,2,2\}\cup \{1,2,4\}|}\\=\frac{|\{1,2\}|}{|\{1,2,2,4\}|}=\frac{3+4}{3+4+4+1}\approx0.58
	\label{eq:jaccard_t1}
\end{multline}
\begin{multline}
	J_2(T_2,Q)=\frac{|T_2\cap Q|}{|T_2\cup Q|}=\frac{|\{2,2,3\}\cap \{1,2,4\}|}{|\{2,2,3\}\cup \{1,2,4\}|}\\=\frac{|\{2\}|}{|\{1,2,2,3,4\}|}=\frac{4}{3+4+4+2+1}\approx0.29
	\label{eq:jaccard_t2}
\end{multline}
We observe a \textit{path jaccard} of $0.58$ for $T_1/Q$ (Equ.~\ref{eq:jaccard_t1}) and $0.28$ for $T_2/Q$ (Equ.~\ref{eq:jaccard_t2}).
Therefore, we translate query path position \textit{query:3} to the target path position \textit{target:3}.

@@

When we sort a graph, we switch the node IDs of the nodes according to the result of the sorting algorithm. For example, if a random sort was applied, all existing node IDs would be replaced with new, random ones (the largest node ID would still correspond to the number of nodes on the graph). We would update the edge and path information as described in the paragraph above. The reordering of nodes has a great influence on how the pangenome looks like (Fig.~\ref{fig:sorting}).

>>>>

\FIXME{Update the references to the sections.}
\begin{itemize}
\item~\textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
\item~\textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
\item~\textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item~\textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item~\textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:extract}).
\item~\textit{odgi explode} breaks the graph into connected components (\S\ref{sec:extract}).
\item~\textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:extract}).
\item~\textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
\item~\textit{odgi unchop} combines unitig nodes (\S\ref{sec:edit}).
\item~\textit{odgi break} removes cycles in the graph (\S\ref{sec:edit}).
\item~\textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
\item~\textit{odgi groom} resolves spurious inverting links (\S\ref{sec:edit}).
\item~\textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).%
%\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
\item~\textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item~\textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
\item~\textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item~\textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item~\textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:supp_metrics}).
\item~\textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:supp_metrics}).
\item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:supp_metrics}).
\item~\textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item~\textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:supp_metrics}).
\item~\textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item~\textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

@@

The ODGI data structure (Algorithm~\ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA file.
Then, we serially add edges in the $Node.edges$ vectors of pairs of nodes.
Finally, we create paths in serial, and extend them in parallel by obtaining the mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
This parallelism speeds ODGI model construction by many-fold when testing against graphs made from assemblies produced by the HPRC (Fig.~\ref{fig:evaluation}).

@@

As a use case study (Fig.~\ref{fig:metrics}), we took a look at the metrics of a 90-haplotype human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}). In particular, we obtained the number of nodes, edges, paths, components, bases, the graph length, and the GC content with \textit{odgi stats}. The output pangenome statistics in YAML textual file format was given to
MultiQC's~\citep{Ewels_2016} newly added ODGI module. As can be seen in Figure~\ref{fig:metrics-multiqc}, we observe a very high GC content of 73.0\% in the \textit{HTTexon1} graph compared to the human genomic mean GC content of 40.9\% \citep{Piovesan2019}. This is in accordance with the literature~\citep{Neueder2017}. %\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)}
Despite this discovery, the MultiQC module provides an interactive way to comparatively explore statistics of an arbitrary number of graphs.

Investigating in detail which intricate regions in the \textit{HTTexon1} graph are responsible for the genetic variation and high GC content, we took a look at the per nucleotide node degree distribution (Fig.~\ref{fig:metrics-degree}), and node depth distribution (Fig.~\ref{fig:metrics-depth}) of CHM13 using \textit{odgi depth}'s, and \textit{odgi degree}'s BED output, respectively. The results indicate a highly polymorphic region around position 200 in the graph. Figure~\ref{fig:metrics-viz} supports this analysis. Zooming in on this region with \textit{vg viz}, we can clearly identify the typical \textit{HTTexon1} CAG variable repeat region (Fig.~\ref{fig:metrics-str}).
Figures~\ref{fig:metrics-degree}-\ref{fig:metrics-viz} highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \citep{Nance1999}.
In general, such complex motives can be the mirror of genetic variation, but also misassemblies or problems in the pangenome building, making the tools further useful for graph validation.
%In Fig.~\ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene~\citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}) are shown.

@@

In the following an example which is based on the graph in Figure~\ref{fig:jaccard}.
\input{fig_jaccard}
Let's assume, we want to translate the query path position \textit{query:3} to a path position in path \textit{target} with a context distance of 2 nucleotides per direction. Node \textit{2} is where \textit{query:3} is located. The node is traversed two times by \textit{target}. Therefore, we apply the \textit{path jaccard} concept to do a precise position translation. Starting from node \textit{2}, we collect nodes $\{1,2,4\}$ for path \textit{query}. For path \textit{target} we have two possible steps to start our collection from: The first or the second step in node \textit{2}. We obtain the node sets $T_1=\{1,2,2\}$ and $T_2=\{2,2,3\}$.
\begin{multline}
	J_1(T_1,Q)=\frac{|T_1\cap Q|}{|T_1\cup Q|}=\frac{|\{1,2,2\}\cap \{1,2,4\}|}{|\{1,2,2\}\cup \{1,2,4\}|}\\=\frac{|\{1,2\}|}{|\{1,2,2,4\}|}=\frac{3+4}{3+4+4+1}\approx0.58
	\label{eq:jaccard_t1}
\end{multline}
\begin{multline}
	J_2(T_2,Q)=\frac{|T_2\cap Q|}{|T_2\cup Q|}=\frac{|\{2,2,3\}\cap \{1,2,4\}|}{|\{2,2,3\}\cup \{1,2,4\}|}\\=\frac{|\{2\}|}{|\{1,2,2,3,4\}|}=\frac{4}{3+4+4+2+1}\approx0.29
	\label{eq:jaccard_t2}
\end{multline}
We observe a \textit{path jaccard} of $0.58$ for $T_1/Q$ (Equ.~\ref{eq:jaccard_t1}) and $0.29$ for $T_2/Q$ (Equ.~\ref{eq:jaccard_t2}).
Therefore, we translate query path position \textit{query:3} to the target path position \textit{target:3}.

@@

When we sort a graph, we switch the node IDs of the nodes according to the result of the sorting algorithm. For example, if a random sort was applied, all existing node IDs would be replaced with new, random ones (the largest node ID would still correspond to the number of nodes on the graph). We would update the edge and path information as described in the paragraph above. The reordering of nodes has a great influence on how the pangenome looks like (Fig.~\ref{fig:sorting}).
\input{fig_sorting}

\subsection{Graph Metrics}
\label{sec:supp_metrics}
Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
To investigate on the \REVIEWED{[1,2] genomic sequences [1,2]} encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
\textit{odgi flatten} generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
ODGI also offers more advanced tools for the interrogation of the graphs.
To study very large pangenomes, users can use \textit{odgi bin} to summarize the path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
This concept found its way into the pangenome graph ontology \citep{Yokoyama2020}. %maybe we want to save this up for the discussion
